Where Do Events Come From?

### 2018-08-15 paper.tex
(>) Revisões Sztajnberg até 2.1

\CEU~\cite{ceu.sensys13} is a Esterel-based~\cite{esterel.ieee91}
%~\cite{ceu.tecs17}
reactive
programming language targeting resource-constrained embedded systems, with the
characteristics that follow:
%
\begin{description}
\item [Reactive:] code only executes in reactions to events and is idle most of
    the time.
\item [Structured:] programs use lexically-scoped structured control
    mechanisms such as \code{spawn} and \code{await} (to create and suspend
    lines of execution).
\item [Synchronous:] reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{description}

Structured reactive programming lets developers write code in direct style,
recovering from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.

@@

Listing~\ref{lst.ceu} illustrates the main characteristics of \CEU, namely
event-driven I/O, lexically-scoped compositions, and synchronous execution.
%
The program toggles the state of the LED whenever a radio packet is received,
terminating on a button press, always with the LED off.
%
The program first declares the \code{LED} output, and the \code{BUTTON} and
\code{RADIO\_RECV} input events (ln~1--3).
The declarations include a payload type, i.e., each event occurrence carries
a value of that type (\code{high/low} is a boolean type).
%
Then, the program uses a \code{par/or} composition (ln~4--16) to run two lines
of execution, aka \emph{trails}, in parallel:
a single-statement trail that waits for a button press before terminating
(ln~5), and an endless loop that toggles the LED on and off whenever a radio
packet is received (ln~10--15).
The \code{finalize} clause (ln~7--9) ensures that, no matter how its enclosing
trail terminates (e.g., from a button press), the LED is unconditionally
turned off~(ln~8).

@@

The synchronous execution model of \CEU dictates that reactions to input events
are atomic and that incoming events are never lost, which we refer to as the
\emph{atomicity} and \emph{responsiveness} properties, respectively.
%
%The event loop in Figure~\ref{fig.env} suggests this behavior since the
%environment can only generate a new input after the application yields control
%and closes the event loop.
%
In Listing~\ref{lst.ceu}, even if two packets arrive simultaneously in the
environment, the synchronous model ensures these properties by adopting an
event queue:
    the first \code{await} awakes and turns the LED off atomically
    (ln~12--13);
    only after that, the second \code{await} will awake (ln~14).
%
%Note that if multiple trails in parallel await the same input, they will all
%awake from it during the same reaction.
%In this case, \CEU dictates that each trail executes atomically, one after the
%other, in lexical order.
%This rule prevents race conditions for trails in parallel.

@@

As a first example, Listing~\ref{lst.gpio.sync}~and~\ref{lst.gpio.async} uses
GPIOs to connect a button to an LED via software such that the LED (pin 13) is
on whenever the button (pin 02) is pressed, and off whenever it is unpressed.

Listing~\ref{lst.gpio.sync}, the synchronous side, first declares its interface
with the external world and includes the asynchronous side as a driver
(ln~1--3).
It then starts with the LED off (ln~5) and enters a loop (ln~6--9) that,
whenever the button changes (ln~7), toggles the state of the LED with the new
value (ln~8).

@@

The example illustrates the clear separation between the application and the
driver through an event-driven interface. % that reflects the desired architecture
%of Figures~\ref{fig.env}~and~\ref{fig.async}.
%
Now, the whole application is written in \CEU:
    the synchronous side remains well behaved with no low-level calls, and
    the asynchronous side deals with the complexity of device drivers.
%
Nevertheless, drivers are typically write-once components developed by embedded
specialists, which are reused in most applications.
%
Note that each supported architecture requires a mapping between the
\code{async/isr} and the actual interrupt vector table (which we provide
for the \emph{AVR/ATmega} and \emph{ARM/Cortex} microcontrollers).

@@

  do finalize with
    {
      ADCSRA &= B01111111;  // disables A/D converter
      ACSR    = B10000000;  // disables comparator
      DIDR0  |= B00111111;  // disables pins
    }
  end
  ...     // driver initialization
  ...     // input / output implementations

@@

\newpage
\subsection{Safe Shared-Memory Concurrency}
\label{sec.ceu.shared}

@@

\newpage
\subsection{Energy-Efficient Runtime}
\label{sec.energy}

>>>>

\CEU~\cite{ceu.sensys13} is an Esterel-based~\cite{esterel.ieee91}
%~\cite{ceu.tecs17}
reactive
programming language targeting resource-constrained embedded systems, with the
characteristics that follow:
%
\begin{description}
\item [Reactive:] code only executes in reaction to events and is idle most of
    the time.
\item [Structured:] programs use lexically-scoped structured control
    mechanisms such as \code{spawn} and \code{await} (to create and suspend
    lines of execution).
\item [Synchronous:] reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{description}

Structured reactive programming lets developers write code in direct style,
reversing from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.

@@

Listing~\ref{lst.ceu} illustrates the main characteristics of \CEU, namely
event-driven I/O, lexically-scoped compositions, and synchronous execution.
%
The program toggles the state of the LED whenever a radio packet is received,
terminating on a button press, always with the LED off.
%
The program first declares the \code{LED} output, and the \code{BUTTON} and
\code{RADIO\_RECV} input events (ln~1--3).
The declarations include a payload type, i.e., each event occurrence carries
a value of that type (\code{high/low} is a boolean type).
%
Then, the program uses a \code{par/or} composition (ln~4--16) to run two lines
of execution, aka \emph{trails}, in parallel:
a single-statement trail that waits for a button press before terminating
(ln~5), and an endless loop that toggles the LED on and off whenever a radio
packet is received (ln~10--15).
The \code{finalize} clause (ln~7--9) ensures that, no matter how its enclosing
trail terminates or aborts (i.e., even from a button press), the LED is
unconditionally turned off~(ln~8).

@@

% AS: Agora entendi ... mas não é muito "interweaved" ou seja ... um trail
% afeta o outro, e voce tira vantagem disso para que a clasula finalize seja
% afetada ... não seria o caso de deixar claro que a clausula finalize está
% fora de qalquer trail? Ou por outra, seu eu tivesse vários trails, todos
% seriam interrompidos. E se cada um tivesse um finalize? todos seriam
% executados? em que ordem?
% FS: A questão é justamente ser menos "interweaved": Se eu escrevi o
% comportamento do LED em uma trilha, essa mesma trilha tem que ser responsável
% por terminar em um estado seguro para o LED (ou qq recurso). Assim, você não
% precisa se preocupar em como as outras trilhas executam ou terminam, o seu
% estado final só depende de você.
% Se eu colocar para fora de qualquer trilha, eu vou ter que misturar o código
% com diversos finalizes que não estão relacionados. Ainda tem o problema de
% saber *onde* colocar. Esse exemplo só tem um par/or, então é óbivo, mas tem
% programas com loops, par/or, par/and, etc aninhados.
% Se cada trilha tiver um finalize e todas forem abortadas, elas executam em
% ordem léxica, que é a mesma regra para quando trilhas acordam na mesma
% reação.

The synchronous execution model of \CEU dictates that reactions to input events
are atomic and that incoming events are never lost, which we refer to as the
\emph{atomicity} and \emph{responsiveness} properties, respectively.
%
%The event loop in Figure~\ref{fig.env} suggests this behavior since the
%environment can only generate a new input after the application yields control
%and closes the event loop.
%
In Listing~\ref{lst.ceu}, even if two packets arrive simultaneously in the
environment, the synchronous model ensures these properties by adopting an
event queue:
    the first \code{await} awakes from the first packet and turns the LED off
    atomically (ln~12--13);
    only after that, the second \code{await} will awake from the second packet
    (ln~14).
From the point of view of the program, the packets arrive in sequence and each
is handled in separate.
%
%Note that if multiple trails in parallel await the same input, they will all
%awake from it during the same reaction.
%In this case, \CEU dictates that each trail executes atomically, one after the
%other, in lexical order.
%This rule prevents race conditions for trails in parallel.

@@

As a first example, Listing~\ref{lst.gpio.sync}~and~\ref{lst.gpio.async} use
GPIOs to connect a button to an LED via software such that the LED (pin 13) is
on whenever the button (pin 02) is pressed, and off whenever it is unpressed.
%
The idea is to use Listing~\ref{lst.gpio.sync} handle synchronous/atomic
changes in the button to affect the LED, and Listing~\ref{lst.gpio.async} to
handle asynchronous changes that are enqueued towards the synchronous side.

Listing~\ref{lst.gpio.sync}, the synchronous side, first declares its interface
with the external world and includes the asynchronous side as a driver
(ln~1--3).
It then starts with the LED off (ln~5) and enters a loop (ln~6--9) that,
whenever the button changes (ln~7), toggles the state of the LED with the new
value (ln~8).
%
This code should be concise and straightforward: each button change toggles the
LED (atomicity) and no button changes are ever lost (responsiveness).

@@

The example illustrates the clear separation between the application and the
driver through an event-driven interface. % that reflects the desired architecture
%of Figures~\ref{fig.env}~and~\ref{fig.async}.
%
Now, the whole application is written in \CEU:
    the synchronous side remains well behaved with no low-level calls, and
    the asynchronous side deals with the complexity of device drivers.
%
Nevertheless, drivers are typically write-once components developed by embedded
systems specialists, which are reused in most applications.
%
Note that each supported architecture requires a mapping between the
\code{async/isr} and the actual interrupt vector table (which we provide
for the \emph{AVR/ATmega} and \emph{ARM/Cortex} microcontrollers).

@@

  finalize with
    {
      ADCSRA &= B01111111;  // disables A/D converter
      ACSR    = B10000000;  // disables comparator
      DIDR0  |= B00111111;  // disables pins
    }
  end
  ...     // driver initialization
  ...     // input / output implementations

@@

\subsection{Safe Shared-Memory Concurrency}
\label{sec.ceu.shared}

@@

\subsection{Energy-Efficient Runtime}
\label{sec.energy}
